---
title: "Analysis"
output: html_document
---

```{r}
library(tidyverse)
library(ggplot2)
marathons <- read_rds('majorMarathons.Rda')
glimpse(marathons)
```

```{r cars}
glimpse(marathons)
```

## Including Plots

You can also embed plots, for example:

```{r, echo=FALSE}
marathons %>%
  filter(!is.na(country)) %>%
  group_by(country)%>%
  summarise(count = n()) %>%
  filter(count>10) %>%
  ggplot(aes(country,count)) + 
    geom_bar(stat="identity",colour="Black",fill="Blue") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    xlab("Country") + 
    ylab("Frequency") +
    ggtitle("Total Major Marathon Wins by Country")
```    
 
```{r}
 marathons %>%
  filter(!is.na(country)) %>%
  group_by(winner,country,gender)%>%
  summarise(count = n()) %>%
  arrange(desc(count))
```
US not a major marathon power, but bill rodgers most marathon major wins 
both top males = US + many wins @ boston (give atble)

top 2 females = norwat (waitz = 1st female < 2:30:00)

```{r pressure, echo=FALSE}
marathons %>%
  filter(!is.na(country),
         year>=1966) %>%
  group_by(country,gender) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  filter(count>5) %>%
  ggplot(aes(country,count)) + 
    geom_bar(aes(fill=gender),stat="identity",colour="Black") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    xlab("Country") + 
    ylab("Frequency") +
    ggtitle("Total Major Marathon Wins by Country")
```  

```{r}
 marathons %>%
  filter(!is.na(country),
         year>=1966) %>%
  group_by(marathon,gender)%>%
  summarise(count = n()) %>%
  arrange(desc(count))

#marathons %>%
#  filter(!is.na(country),
#         year>=1966) %>%
  #group_by(country,gender) %>%
  #summarise(count = n()) %>%
  #arrange(desc(count)) %>%
  #filter(count>5) %>%
  ggplot(marathons,aes(x=country)) + 
    geom_bar() + 
    coord_cartesian(ylim=c(30,115))
```
  aes(y = ..count.., fill=gender),stat="identity",colour="Black",
             position="dodge") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    xlab("Country") + 
    ylab("Frequency") +
    ggtitle("Total Major Marathon Wins by Country") + 
  facet_grid(.~marathon)


library(data.table)
mdt <- marathons %>%
  mutate(year=as.integer(year)) %>%
  as.data.table()
dcast.data.table(mdt, country ~ year,
                           fun.aggregate = count)

  
  
```{r}
filter(marathons,country=="Kenya")[which.min(filter(marathons,country=="Kenya")$year),]
```
```{r}
# wins bykenya by year
marathons %>%
  filter(country %in% c("Kenya","Norway","Unites States","United Kingdom","Germany","Ethiopia","West Germany","Japan")) %>%
  group_by(country,year) %>%
  summarize(count = n()) %>%
  ggplot(aes(x=year,y=cumsum(count),group=country,color=country)) + 
    geom_freqpoly(stat = "identity") + 
  coord_cartesian(xlim=c(1985,2018))
#length(marathons$year)
```

```{r}
marathons %>% 
  group_by(country, year) %>%
  summarise(value = n()) %>%
  mutate(csum = cumsum(value)) %>%
  group_by(country) %>%
  summarise(n = max(csum)) %>%
  arrange(desc(n))
```

```{r, warning=F,message=FALSE}
marathons %>% 
  group_by(country, year) %>%
  summarise(value = n()) %>%
  #mutate(csum = cumsum(value)) %>%
  filter(country %in% c("Kenya","Norway","United States","United Kingdom","Germany","Ethiopia","West Germany","Japan")) %>%
  ggplot(aes(x=year,color=country)) +#,group=country,col=country)) +
  stat_bin(data=subset(marathons,country=="Kenya"),aes(y=cumsum(..count..)),geom="line")+
  stat_bin(data=subset(marathons,country=="Norway"),aes(y=cumsum(..count..)),geom="line")+
  stat_bin(data=subset(marathons,country=="Germany"),aes(y=cumsum(..count..)),geom="line")+
  stat_bin(data=subset(marathons,country=="West Germany"),aes(y=cumsum(..count..)),geom="line")+
  stat_bin(data=subset(marathons,country=="United Kingdom"),aes(y=cumsum(..count..)),geom="line")+
  stat_bin(data=subset(marathons,country=="Japan"),aes(y=cumsum(..count..)),geom="line")+
  stat_bin(data=subset(marathons,country=="United States"),aes(y=cumsum(..count..)),geom="line")
```
compare wiht olympics + world champs?